---
title: "Final project"
author: "Daniel Gonzalez-Suarez"
date: "4/3/2020"
output: 
  pdf_document: 
    toc: yes
  html_document: 
    theme: united
    toc: yes
editor options:
  chunk_output_style: console
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part II, Final Assignment
This is the second assingment for the Ecological Data Analysis in R course. In order to start, packages and functions will be loaded first.
```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(vegan)
library(viridis) #just in case. For ggplot, use scale_color_viridis() and scale_fill_viridis() magma, plasma, inferno, viridis
source("./functions/6003Functions.R")
```

## Loading data and tidying it up.

Here, steps for tidying the dataset will be presented. Raw data was taken during completion of underwater transects, then input in Excel. Data will be imported as a .csv file.  

```{r}
df <- read.csv("./data/df_daniel.csv")


#df contains data for analysis
head(df)
tail(df)
str(df)
range(df$abundance)


# ab_matrix contains abundance matrix for ecological indices
# we'll create it out of df; as we don't care for sizes, we'll sum all abundances for each species on for province, site and transect
# first, we'll summarise species disregarding length
# we'll use two different matrices: one for species and one for taxa
# not all organisms were identified to species level
# taxa for invertebrates is ranked down to class, subclass or order. for fish, taxa equals family.

df_sp <- df %>% select(province, locality, transect, vert_invert, taxa, sp, sp_code, abundance)
head(df_sp)
str(df_sp)
df_sp <- summarise(group_by(df_sp, province, locality, transect, sp, sp_code),
                        sum_abundance = sum(abundance, na.rm = TRUE))

df_taxa <- df %>% select(province, locality, transect, vert_invert, taxa, abundance)
head(df_taxa)
str(df_taxa)
df_taxa <- summarise(group_by(df_taxa, province, locality, transect, taxa),
                        sum_abundance = sum(abundance, na.rm = TRUE))


# let's set up the data so that we can create our matrices
preab_sp  <- pivot_wider(df_sp, names_from = sp, values_from = sum_abundance)
preab_sp <- column_to_rownames(preab_sp, var="transect")
preab_taxa  <- pivot_wider(df_taxa, names_from = taxa, values_from = sum_abundance)

ab_sp <- preab_sp[, 5:121] 

ab_taxa <- preab_taxa[, 5:50]

# not sure if necessary but let's convert na to zeroes
#ab_sp[is.na(ab_sp)] <-0 
#ab_taxa[is.na(ab_taxa)] <-0 


#let's just take a look at our data before going on
boxplot(df_sp$sum_abundance)
boxplot(df_taxa$sum_abundance)

dotchart(df_sp$sum_abundance)
dotchart(df_taxa$sum_abundance)


```

Everything looks good. Data was already in long format before importing it. 



## Indices

```{r}

#Let's create a dataframe for our indices.
#We need to use only abundance

ind_sp <- preab_sp[,c("province","locality","transect")]
ind_taxa <- preab_taxa[,c("province","locality","transect")]

# Richness: number of species per transect/habitat
ind_sp$Richness <- rowSums(ab_sp>0)
ind_taxa$Richness <- rowSums(ab_taxa>0)

# Shannon´s diversity index: the bigger, the more diverse, basically 
ind_sp$Shannon <- diversity(ab_sp) 
ind_taxa$Shannon <- diversity(ab_taxa)

# Rarefaction
ind_sp$Rarefied <- c(rarefy(ab_sp[1:1170,], sample=15))
ind_taxa$Rarefied <- c(rarefy(ab_taxa[1:46,], sample=15))

# Also visualize a rarefaction curve
rarecurve(ab_sp)

```


