---
title: "Final project"
author: "Daniel Gonzalez-Suarez"
date: "4/3/2020"
output: 
  pdf_document: 
    toc: yes
  html_document: 
    theme: united
    toc: yes
editor options:
  chunk_output_style: console
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Assignment
This is the second assingment for the Ecological Data Analysis in R course. In order to start, packages and functions will be loaded first.
We'll use _tidiverse_, _dplyr_, _vegan_, _viridis_ and a customized function that will be available in the repository.
```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(vegan)
library(viridis)
#source("./functions/6003Functions.R")
source("D:/UABC/II_Semestre/Datos_R/eda_finalproject/functions/6003Functions.R") #this one's for knitting of document
```

## Loading data and tidying it up.

Here, steps for tidying the dataset will be presented. Raw data was taken during completion of underwater transects, then input in Excel. Data will be imported as a .csv file.  

```{r}
#df <- read.csv("./data/df_daniel.csv")
df <- read.csv("D:/UABC/II_Semestre/Datos_R/eda_finalproject/data/df_daniel.csv") #this one's for knitting of document

# We'll only need provinces, localities, sites, transects, species and abundances
df <- df %>% select(province, locality, siteref, transect, sp, length, abundance)

head(df, n=3L)
tail(df, n=3L)
str(df)
```

Before continuing, let's remember our question: will there be a difference between communities of two provinces with different fishing pressures?

## Data Exploration
### 1. Outliers 
```{r}
boxplot(df$length)
boxplot(df$abundance)

dotchart(df$length)
dotchart(df$abundance)
```
Not that uncommon to see abundances represented like that, I think. Lot's of zeroes and some outliers for abundance and no errors regarding length.

### 2. Homogeneity of variance 
```{r}
p <- ggplot(data=df, aes(X=province, y=length)) +
  geom_boxplot() + facet_wrap(~siteref)
p
p<- ggplot(data=df, aes(X=province, y=abundance)) +
  geom_boxplot() + facet_wrap(~siteref)
p
```

### 3. Normality 
```{r, include=FALSE}
hist(df$length)
hist(df$abundance)

qqnorm(df$length)
qqline(df$length)

qqnorm(df$abundance)
qqline(df$abundance)
```

Data is not normal, as seen in histograms and confirmed by Q-Q plots.

### 4. Zeroes 
```{r}
range(df$length)
plot(table(df$length))

range(df$abundance)
plot(df$abundance)

```

There are lots of zeroes for abundance, which we'll have to take into account when analyzing data.

### 5. Collinearity X
```{r}
pairs(~ province + locality + siteref + transect + length + abundance, 
      lower.panel=panel.smooth, upper.panel=panel.cor, 
      data=df) 

plot(length ~ province, data=df)
plot(length ~ locality, data=df)
plot(length ~ siteref, data=df)
plot(length ~ transect, data=df)

plot(abundance ~ province, data=df)
plot(abundance ~ locality, data=df)
plot(abundance ~ siteref, data=df)
plot(length ~ transect, data=df)

pairs(~ province+locality+siteref+transect+abundance, 
      lower.panel=panel.smooth, upper.panel=panel.cor, 
      data=df) 

```

We'll take into account province, locality, site and transects as x variables and abundance as y variables; our question does not requires us to study length, so it's something we could have dropped before. We'll still explore it for the sake of it.

### 6. Relationships X and Y
```{r}
p <- ggplot(df, aes(x=province, y=length)) +
  geom_boxplot()
p

p <- ggplot(df, aes(x=province, y=abundance)) +
  geom_boxplot()
p
```


### 7. Interactions
```{r}
coplot(length ~ abundance | province ,
       data=df,
       panel=function(x,y,...) {
         panel.smooth(x,y,span=0.8,iter=5,...)
         abline(lm(y ~ x), col="blue") } )

coplot(length ~ abundance | locality ,
       data=df,
       panel=function(x,y,...) {
         panel.smooth(x,y,span=0.8,iter=5,...)
         abline(lm(y ~ x), col="blue") } )

coplot(length ~ abundance | siteref ,
       data=df,
       panel=function(x,y,...) {
         panel.smooth(x,y,span=0.8,iter=5,...)
         abline(lm(y ~ x), col="blue") } )

#coplot(length ~ abundance | transect ,
#       data=df,
#       panel=function(x,y,...) {
#         panel.smooth(x,y,span=0.8,iter=5,...)
#         abline(lm(y ~ x), col="blue") } )

```

There seems to be no interaction. 

### 8. Independence of Y
Both length and density could be dependent on province, locality, site because of fishing pressure effects. In transects, I'd say there could be independence.

## Indices
### Data setup
We'll need to create an abundance matrix in order to calculate indices.
Let's create a dataframe for our indices.

```{r}
#Let's create a dataframe for our indices.
#We only need abundance

# first, we'll summarise species disregarding length 
df_sp <- df %>% select(province, locality, siteref, transect, sp, abundance)
head(df_sp)
str(df_sp)
df_sp <- summarise(group_by(df_sp, province, locality, siteref, transect, sp),
                        sum_abundance = sum(abundance, na.rm = TRUE))

# ab_sp contains abundance matrix for ecological indices
# we'll create it out of df; as we don't care for sizes, we'll sum all abundances for each species on for province, site and transect
ab_sp  <- pivot_wider(df_sp, names_from = sp, values_from = sum_abundance)
#converting these factors to numeric will solve an issue that could arise later when we're creating our indices
ab_sp$province <- as.double(ab_sp$province) 
ab_sp$locality <- as.double(ab_sp$locality)
ab_sp$siteref <-as.double(ab_sp$siteref)
ab_sp$transect <- as.double(ab_sp$transect)

# let's convert na to zeroes
ab_sp[is.na(ab_sp)] <-0 

```

Now we can obtain ecological indices
```{r}
#We'll create a dataframe to store our results
ind_sp <- ab_sp[, c("province", "locality", "siteref", "transect")]

# Richness: number of species per transect/habitat
ind_sp$richness <- rowSums(ab_sp>0)

# Shannon´s diversity index: the bigger, the more diverse, basically 
ind_sp$Shannon <- diversity(ab_sp) 

# Rarefaction
raremax <- min(rowSums(ab_sp>0))
ind_sp$Rarefied <- c(rarefy(ab_sp[1:35,], sample=raremax))
raremax <- min(rowSums(ab_sp>0))

# Let's also visualize a rarefaction curve
#this function will create a rarefaction curve to observe species accumulation 
rarecurve(ab_sp, sample = raremax, col=viridis(raremax, alpha = 1, begin = 0, end = 1, direction = 1, option = "D"), lwd=2.7)
```

### Visualization
We'll use boxplots to see  differences in species diversity:

```{r}
#let's return our province/location/sites to factors
par(mfrow=c(1,2))
boxplot(richness~province, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Richness")
boxplot(richness~siteref, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Richness")
boxplot(Shannon~province, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Shannon diversity")
boxplot(Shannon~siteref, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Shannon diversity")
boxplot(Rarefied~province, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Rarefied richness")
boxplot(Rarefied~siteref, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Rarefied richness")

mfrow=c(1,1)
```

Looks like there are "some" differences, so let's explore them.

### Linear modeling of indices

It appears that species diversity increases as we move from the field to the forest. We can test for differences among habitats statistically using a linear model, with Habitat as a predictor of species diversity:

```{r}
# fit linear models (ANOVA)
mod.richness.province <- lm(richness~province, data=ind_sp)
mod.richness.siteref <- lm(richness~siteref, data=ind_sp)
mod.richness.transect <- lm(richness~transect, data=ind_sp)

mod.Shannon.province <- lm(Shannon~province, data=ind_sp)
mod.Shannon.siteref <- lm(Shannon~siteref, data=ind_sp)
mod.Shannon.transect <- lm(Shannon~transect, data=ind_sp)

mod.Rarefied.province <- lm(Rarefied~province, data=ind_sp)
mod.Rarefied.siteref <- lm(Rarefied~siteref, data=ind_sp)
mod.Rarefied.transect <- lm(Rarefied~transect, data=ind_sp)

anova(mod.richness.province)
anova(mod.richness.siteref) 
anova(mod.richness.transect)
#Results show no significant differences

anova(mod.Shannon.province)
anova(mod.Shannon.siteref) 
anova(mod.Shannon.transect)
#There are significant differences for each one 

anova(mod.Rarefied.province)
anova(mod.Rarefied.siteref)
anova(mod.Rarefied.transect)
#Same!

#Should we compare between groups? Let's try a Tukey test. Also, I'm really not sure what am I doing but I think I'm going in the right direction... right?
#Tukey for Shannon
TukeyHSD(aov(lm(Shannon~as.factor(siteref), data = ind_sp)), conf.level = 0.95)

#Differences for  site1-4, site1-5, site2-4, site2-5 and almost site5-6

#Tukey for Rarefied richness
TukeyHSD(aov(lm(Rarefied~as.factor(siteref), data = ind_sp)), conf.level = 0.95)

#Differeces for site 1-4, site1-5, site2-4, site2-5, site3-5 and almost site5-6

```

### Interpretation
Results of ANOVA show no significant difference for richness, but does show it for Shannon and rarefied species richness. Seems like there's no association between province/site and species richness, but it is for Shannon's diversity index and rarefied richness.

Should we go on? Yes! Well... Seems like my data won't allow me to continue, if I'm not mistaken. I was considering doing an ordination analysis to determine which sites were similar and what species share distribution in provinces, sites and/or transects. I'd need to dampen data from abundances and divide it by ...? I don't know. So I'll leave it up to here. 

Based on what I have, the answer to my question would be that there's evidence -based on my analysis- that differences in species composition could be explained by the location, site or transect. Based on Tukey test, there are some differences between provinces: sites 1-3 correspond to Puntarenas, the high fishing pressure province, and 4-7 to Guanacaste, the not-so-high fishing pressure province. 

## Acknowledgement
Luis, I'd like to thank you for your effort, patience and dedication given to this course and us. I'm deeply grateful I got into it. I've learned a lot and have also improved some practices that I didn't know were not so great. Data exploration has been (and will continue to be) really useful to me, and will be fundamental for each research project. Thank you, Luis! You're a great professor and a remarkable human being, even if we have only shared a few months. 