---
title: "Final project"
author: "Daniel Gonzalez-Suarez"
date: "4/3/2020"
output: 
  pdf_document: 
    toc: yes
  html_document: 
    theme: united
    toc: yes
editor options:
  chunk_output_style: console
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part II, Final Assignment
This is the second assingment for the Ecological Data Analysis in R course. In order to start, packages and functions will be loaded first.
```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(vegan)
library(viridis) #just in case For ggplot, use scale_color_viridis() and scale_fill_viridis() magma, plasma, inferno, viridis
source("./functions/6003Functions.R")
```

## Loading data and tidying it up.

Here, steps for tidying the dataset will be presented. Raw data was taken during completion of underwater transects, then input in Excel. Data will be imported as a .csv file.  

```{r}
df <- read.csv("./data/df_daniel.csv")
# We'll only need provinces, localities, transects, species, length and abundances
df <- df %>% select(province, locality, transect, sp, length, abundance)

head(df, n=3L)
tail(df, n=3L)
str(df)

```

Everything looks good. Data was already in long format before importing it. 


## Exploring the data
We are going to see visual representation of data.

# 1. Outliers 

```{r, include=FALSE}
boxplot(df$length)
boxplot(df$abundance)
boxplot(df$Densidad_.org.m2.)

dotchart(df$Talla_.cm.)
dotchart(df$Abundancia)
dotchart(df$Densidad_.org.m2.)
```

Many outliers, but there are no errors on data.

# 2. Homogeneity of variance 

```{r, include=FALSE}
p <- ggplot(data=df, aes(X=Provincia, y=Talla_.cm.)) +
  geom_boxplot() + facet_wrap(~SitioCodigo)
p
p<- ggplot(data=df, aes(X=SitioCodigo, y=Densidad_.org.m2.)) +
  geom_boxplot() + facet_wrap(~Localidad)
p
```

There is homogeneity of variance.

# 3. Normality 

```{r, include=FALSE}
hist(df$length)
hist(df$abundance)

qqnorm(df$length)
qqline(df$abundance)

qqnorm(df$Abundancia)
qqline(df$Abundancia)

qqnorm(df$Densidad_.org.m2.)
qqline(df$Densidad_.org.m2.)

```

Data is not normal, as seen in histograms and confirmed by Q-Q plots.

# 4. Zeroes 

```{r}
range(df$Talla_.cm.)
plot(table(df$Talla_.cm.))

range(df$Abundancia)
plot(df$Abundancia)

range(df$Densidad_.org.m2.)
plot(table(df$Densidad_.org.m2.))

```

There are no zeroes in either continuous variable.


# 5. Collinearity X
```{r}
pairs(~ Provincia + Localidad + SitioCodigo + Densidad_.org.m2. + Talla_.cm., 
      lower.panel=panel.smooth, upper.panel=panel.cor, 
      data=df) 

plot(Talla_.cm. ~ Provincia, data=df)
plot(Talla_.cm. ~ Localidad, data=df)
plot(Talla_.cm. ~ SitioCodigo, data=df)

plot(Densidad_.org.m2. ~ Provincia, data=df)
plot(Densidad_.org.m2. ~ Localidad, data=df)
plot(Densidad_.org.m2. ~ SitioCodigo, data=df)

pairs(~ Provincia + Densidad_.org.m2. + Talla_.cm., 
      lower.panel=panel.smooth, upper.panel=panel.cor, 
      data=df) 
```

Our X variable will now be only Province

# 6. Relationships X and Y
```{r}
p <- ggplot(df, aes(x=Provincia, y=Talla_.cm.)) +
  geom_boxplot()
p

p <- ggplot(df, aes(x=Provincia, y=Densidad_.org.m2.)) +
  geom_boxplot()
p
```


# 7. Interactions

```{r}

coplot(Talla_.cm. ~ Densidad_.org.m2. | Provincia ,
       data=df,
       panel=function(x,y,...) {
         panel.smooth(x,y,span=0.8,iter=5,...)
         abline(lm(y ~ x), col="blue") } )
```
There seems to be no interaction in Provinces. 

# 8. Independence of Y
Both length and density could be dependent on Province because of fishing pressure effects. 

