---
title: "Final project"
author: "Daniel Gonzalez-Suarez"
date: "4/3/2020"
output: 
  pdf_document: 
    toc: yes
  html_document: 
    theme: united
    toc: yes
editor options:
  chunk_output_style: console
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Part II, Final Assignment
This is the second assingment for the Ecological Data Analysis in R course. In order to start, packages and functions will be loaded first.
```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(vegan)
library(viridis)
source("./functions/6003Functions.R")
```

## Loading data and tidying it up.

Here, steps for tidying the dataset will be presented. Raw data was taken during completion of underwater transects, then input in Excel. Data will be imported as a .csv file.  

```{r}
df <- read.csv("./data/df_daniel.csv")
# We'll only need provinces, localities, sites, transects, species and abundances
df <- df %>% select(province, locality, siteref, transect, sp, length, abundance)

head(df, n=3L)
tail(df, n=3L)
str(df)

```

#Before continuing, let's remember our question: will there be a difference between communities of two provinces with different fishing pressures?

## Data Exploration
# 1. Outliers 
```{r, include=FALSE}
boxplot(df$length)
boxplot(df$abundance)

dotchart(df$length)
dotchart(df$abundance)
```
Not that uncommon to see abundances represented like that, I think. Lot's of zeroes and some outliers for abundance and no errors regarding length.

# 2. Homogeneity of variance 

```{r}
p <- ggplot(data=df, aes(X=province, y=length)) +
  geom_boxplot() + facet_wrap(~siteref)
p
p<- ggplot(data=df, aes(X=SitioCodigo, y=Densidad_.org.m2.)) +
  geom_boxplot() + facet_wrap(~siteref)
p
```

# 3. Normality 

```{r, include=FALSE}
hist(df$length)
hist(df$abundance)

qqnorm(df$length)
qqline(df$length)

qqnorm(df$abundance)
qqline(df$abundance)
```

Data is not normal, as seen in histograms and confirmed by Q-Q plots.

# 4. Zeroes 

```{r}
range(df$length)
plot(table(df$length))

range(df$abundance)
plot(df$abundance)

```

There are lots of zeroes for abundance, which we'll have to take into account when analyzing data.


# 5. Collinearity X
```{r}
pairs(~ province + locality + siteref + transect + length + abundance, 
      lower.panel=panel.smooth, upper.panel=panel.cor, 
      data=df) 

plot(length ~ province, data=df)
plot(length ~ locality, data=df)
plot(length ~ siteref, data=df)
plot(length ~ transect, data=df)

plot(abundance ~ province, data=df)
plot(abundance ~ locality, data=df)
plot(abundance ~ siteref, data=df)
plot(length ~ transect, data=df)

#We'll try to take only into account province as x variable and length or abundance as y variables
pairs(~ province+ length + abundance, 
      lower.panel=panel.smooth, upper.panel=panel.cor, 
      data=df) 

```

Our X variable will now be province, being length or abundance our y variable

# 6. Relationships X and Y
```{r}
p <- ggplot(df, aes(x=province, y=length)) +
  geom_boxplot()
p

p <- ggplot(df, aes(x=province, y=abundance)) +
  geom_boxplot()
p
```


# 7. Interactions

```{r}

coplot(length ~ abundance | province ,
       data=df,
       panel=function(x,y,...) {
         panel.smooth(x,y,span=0.8,iter=5,...)
         abline(lm(y ~ x), col="blue") } )
```
There seems to be no interaction in Provinces. 

# 8. Independence of Y
Both length and density could be dependent on Province because of fishing pressure effects. 


## Indices
# Data setup
We'll need to create an abundance matrix in order to calculate indices.

```{r}
#Let's create a dataframe for our indices.
#We only need abundance

# first, we'll summarise species disregarding length
# we'll use two different matrices: one for species and one for taxa
# not all organisms were identified to species level
# taxa for invertebrates is ranked down to class, subclass or order. for fish, taxa equals family.
 
df_sp <- df %>% select(province, locality, siteref, transect, sp, abundance)
head(df_sp)
str(df_sp)
df_sp <- summarise(group_by(df_sp, province, locality, siteref, transect, sp),
                        sum_abundance = sum(abundance, na.rm = TRUE))

# ab_sp contains abundance matrix for ecological indices
# we'll create it out of df; as we don't care for sizes, we'll sum all abundances for each species on for province, site and transect
ab_sp  <- pivot_wider(df_sp, names_from = sp, values_from = sum_abundance)
#converting these factors to numeric will solve an issue that could arise later when we're creating our indices
ab_sp$province <- as.double(ab_sp$province) 
ab_sp$locality <- as.double(ab_sp$locality)
ab_sp$siteref <-as.double(ab_sp$siteref)
ab_sp$transect <- as.double(ab_sp$transect)

# let's convert na to zeroes
ab_sp[is.na(ab_sp)] <-0 

```

# Now we can obtain ecological indices
```{r}
#We'll create a dataframe to store our results
ind_sp <- ab_sp[, c("province", "locality", "siteref", "transect")]

# Richness: number of species per transect/habitat
ind_sp$richness <- rowSums(ab_sp>0)

# Shannon´s diversity index: the bigger, the more diverse, basically 
ind_sp$Shannon <- diversity(ab_sp) 

# Rarefaction
ind_sp$Rarefied <- c(rarefy(ab_sp[1:35,], sample=raremax))
raremax <- min(rowSums(ab_sp>0))

# Let's also visualize a rarefaction curve
#this function will create a rarefaction curve to observe species accumulation 
rarecurve(ab_sp, sample = raremax, col=viridis(raremax, alpha = 1, begin = 0, end = 1, direction = 1, option = "D"), lwd=2.7)
```


# Visualization!
We'll use boxplots to see  differences in species diversity:

```{r}
#let's return our province/location/sites to factors
par(mfrow=c(1,2))
boxplot(richness~province, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Richness")
boxplot(richness~siteref, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Richness")
boxplot(Shannon~province, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Shannon diversity")
boxplot(Shannon~siteref, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Shannon diversity")
boxplot(Rarefied~province, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Rarefied richness")
boxplot(Rarefied~siteref, data=ind_sp, boxwex=0.5, col=viridis(35), 
        cex.axis=0.5, ylab="Rarefied richness")

mfrow=c(1,1)
```

## Some linear modeling

It appears that species diversity increases as we move from the field to the forest. We can test for differences among habitats statistically using a linear model, with Habitat as a predictor of species diversity:

```{r}
# fit linear models (ANOVA)
mod.richness.province <- lm(richness~province, data=ind_sp)
mod.richness.siteref <- lm(richness~siteref, data=ind_sp)

mod.Shannon.province <- lm(richness~province, data=ind_sp)
mod.Shannon.siteref <- lm(richness~siteref, data=ind_sp)

mod.Rarefied.province <- lm(richness~province, data=ind_sp)
mod.Rarefied.siteref <- lm(richness~siteref, data=ind_sp)

anova(mod.richness.province)
anova(mod.richness.siteref) 

anova(mod.Shannon.province)
anova(mod.Shannon.siteref) 

anova(mod.Rarefied.province)
anova(mod.Rarefied.siteref)
``